Models for spatial structure in small-area estimation

### 2022-07-13 src/docs_paper/paper.Rmd
Work on manuscript and simulation study

---
title: Models for spatial structure in small-area estimation
author:
  - name: Adam Howes
    email: ath19@ic.ac.uk
    affiliation: A
    corresponding: ath19@ic.ac.uk
  # - name: Harrison Zhu
  #   affiliation: A
  # - name: Yidan Xu
  #   affiliation: C
  # - name: Samir Bhatt
  #   affiliation: D
  - name: Jeff W. Eaton
    affiliation: B
  - name: Seth R. Flaxman
    affiliation: E
address:
  - code: A
    address: Department of Mathematics, Imperial College London
  - code: B
    address: MRC Centre for Global Infectious Disease Analysis, Imperial College London
  # - code: C
  #   address: Department of Statistics, University of Michigan
  # - code: D
  #   address: Department of Public Health, University of Copenhagen
  - code: E
    address: Department of Computer Science, University of Oxford
abstract: |
  **Background:**

@@

@@

Spatial random effects are frequently used to model spatial variation in areal data [@haining2003spatial; @cramb2018investigation].
Gaussian Markov random fields (GMRFs) [@rue2005gaussian], which combine a Gaussian distribution with Markov conditional independence assumptions between areas, are the most common class of models used to specify spatial random effects.
Using a GMRF often reflects an assumption that observations made in areas close together are correlated, and more distant relationships may be ignored.
Arguably the simplest GMRF model is that of @besag1991bayesian, where information is borrowed equally from each adjacent area.
This model requires minimal additional modelling choices and is widely implemented, including as a part of the `R-INLA` package [@rue2017bayesian], popular within spatial statistics.
Use of this model is widespread, including in agriculture [@oliver2015soil], ecology [@saracco2010modeling], epidemiology [@lawson2013statistical], image analysis [@schmid2006bayesian], neuroscience [@gossl2001bayesian] and public health [@dwyer2015drinking].
However, for irregular geometries like the administrative divisions of a country, the assumptions made about space are unrealistic.
We hypothesised that incorporating a spatial structure more faithfully to the specific geometry would improve the performance of small-area estimation models.
We set out to test this hypothesis, and in doing so provide practical recommendations for modelling spatial structure.
<!-- @gutreuter2019improving use the SAR model, which is still adjacency based but is not exactly the Besag model--restructure introduction to be more general and include introduction of other adjacency based models (which all have the same criticisms)? -->

Within the GRMF framework, attempts to more accurately reflect irregular spatial structure have defined weights specifying the extent to which neighbours are related.
@duncan2017spatial compared seventeen methods for specifying these weights, but surprisingly did not find any which outperform the Besag model^[This conclusion was specific to the (often-analysed) Scottish lip cancer example, and like the majority of the literature, based on the deviance information criteria (DIC). The DIC is based on summarising the posterior distribution by a point estimate, and is recommended against by @vehtari2017practical.].
Another approach is to take into account that areal data is typically produced by aggregating higher resolution data, so as to be a particular instance of the broader challenge of inference about a variable at a different resolution to that it was observed at, known in geostatistics as the change-of-support problem [@gelfand2001change] and dating to foundational work on block kriging [@krige1951statistical; @matheron1976forecasting].
Ideas from change-of-support modelling are applied by @kelsall2002modeling to consider areal data as coming about by an aggregation of a continuously-indexed Gaussian process, resulting in a covariance structure between two areas given by the average covariance between two points chosen randomly from those areas.
This type of model is particularly natural in the log-Gaussian Cox process modelling framework [@li2012log; @diggle2013spatial; @taylor2015bayesian; @johnson2019spatially], `lgcp` [@taylor2015bayesian] and `SDALGCP` [@johnson2019spatially] packages.
Inference for aggregated data has recently been advanced by @wilson2018pointless who consider the SPDE approach of @lindgren2011explicit, using `R-INLA`, and an empirical Bayes approach, using `TMB` [@kristensen2015tmb], and made available as a part of the `disaggregation` package [@nandi2020disaggregation].

We also hypothesised that improvements to the Besag model may be transferable to improving convolution models.
Recognising that spatial heterogeneity in outcomes commonly reflects a combination of both spatially correlated processes and specific circumstances at a given location, it is usually recommended to use a random effect specification which includes both spatially structured and unstructured components.
Examples include the BYM2 model [@simpson2017penalising] and earlier convolutions such as the BYM [@besag1991bayesian], Leroux [@leroux2000estimation] or Dean [@dean2001detecting] models.
Evidence for this recommendation includes the comparison studies of @lee2011comparison and @riebler2016intuitive.
These convolutions use a Besag model as the spatially structured component, though any choice of spatial random effect model is possible.
The Naomi HIV small-area estimation model [@eaton2021naomi], which motivates our work, uses both Besag and BYM2 spatial random effects, at different levels of the hierarchical model, and more broadly both models are of substantive practical interest.

@@

The remainder of this paper is organised as follows.
Section \ref{sec:background} provides background on areal data and the Bayesian hierarchical modelling approach to small-area estimation.
In Section \ref{sec:ab}, we review developments in specifying spatial random effects based on adjacency, before presenting an alternative kernel-based approach in Section \ref{sec:ks}.
In Section \ref{sec:simulation} we compare models using simulated data, before applying them to mapping HIV prevalence in Section \ref{sec:hiv}.
Finally, in Section \ref{sec:conclusion} we discuss our conclusions and directions for future research.

@@

Let $\Sc \subset \mathbb{R}^2$ be the study region and the disjoint areas $\{A_i\}_{i = 1}^n$ be a spatial partition of $\Sc$.
Areal data are a type of spatial data where observations $y_i$ are associated to areas $A_i$.
Examples include the colour of a pixel, the minimum wage in a state, and the number of disease cases in a region.
Point data are another type of spatial data where instead observations $y(s)$ can be made at any location $s \in \Sc$ of a spatially-continuous stochastic process.
Often areal may be conceptualised as arising from aggregation of point data so that $y_i = \int_{A_i} y(s) \text{d}s$.
There are some exceptions, such as the minimum wage or disease control measures, which are determined at an administrative level and have a genuinely discrete spatial structure\footnote{For data of this sort, there is less clear justification for using spatial smoothing. Realistically, data relate to many covariates, some number of which are likely have discrete spatial structure and others of which continuous spatial structure, such that the result is often an intermediate. It may be possible to model this explicitly, though the result might look similar to what is achieved using e.g. BYM2.}.

If the areal data we observe are indeed an aggregation of point data, why not use the higher resolution point data in our models rather than areal data?
Explainations include that (a) the underlying point data are unobserved, (b) although point data are collected, they may not be accessible due to privacy constraints, administrative practicality or storage capacity, and (c) the point data are available, but we decide not to use them in the model.
In the final case (c), researchers may prefer to use area-level models because models for point data are typically more complex, require more data, including covariate data which may not exist, and have higher computational burden.
Choosing to model the underlying data generating mechanism as either areal or point is often a matter of pragmatism.

@@

The latter benefit is basis for the statistical task of small-area estimation, which aims to produce reliable local estimates where small sample sizes lead to noisy data [@pfeffermann2013new].
In a spatial setting, this is often in small geographic areas, though the phrase "small-area" is not restricted to geographic areas and may be interpreted more broadly to mean any area where data are insufficient to make accurate local parameter inferences^[For example, in the context of multilevel regression and post-stratification [@gelman1997poststratification] small-areas are generated by the intersection of demographic variables like age, gender and race, alongside geographic variables.].
For example, due to the cost of gathering samples, a survey may be designed to give reliable estimates at a national level but not at a district level.
Although direct estimators of local parameters are unbiased, when data are sparse the total error may be reduced by accepting some bias in exchange for reduced variance using so-called indirect estimators.
Smoothing approaches use information from similar units to "borrow strength" from one parameter to another, with determining precisely what is meant by "similar" a central challenge.
@wakefield2020small review design-based approaches and model-based approaches to spatial analysis of health-indicators, including both area and unit-level analysis.

@@

Bayesian hierarchical models are an attractive framework for small-area estimation [@rao2015small].
To model areal observations $\y$ corresponding to the areas $\{A_i\}_{i = 1}^n$ of a study region $\Sc$, a three-layer structure [@berliner1996hierarchical; @cressie2015statistics] given by
\begin{alignat}{2}
&\text{(Observations)}  &        y_i &\sim p(y_i \, | \, g^{-1}(x_i), \btheta), \quad i = 1, \ldots, n, \label{eq:data} \\
&\text{(Latent field)}  &        \x &\sim p(\x \, | \, \btheta), \label{eq:process} \\
&\text{(Parameters)}    & \qquad \btheta &\sim p(\btheta), \label{eq:parameters}
\end{alignat}
may be used, where $\x$ is the $n$-dimensional latent field, $\btheta$ are the $m$-dimensional\footnote{Ordinarily $m$ is much smaller than $n$ as few parameters are required in order to control the properties of the latent field and likelihood.} parameters and $g$ is a link function.
Spatial structure is modelled at the level of the latent field, allowing the observation model, often either binomial or Poisson, to be conditionally independent and identically distributed $p(\y \, | \, \x, \btheta) = \prod_{i = 1}^n p(y_i \, | \, x_i, \btheta)$.
As small-area estimation models are usually descriptive rather than mechanistic, it is natural [@best2005comparison] to use a Gaussian distribution $\x \sim \mathcal{N}(\bm{\mu}, \mathbf{\Sigma})$ for the latent field^[The Gaussian distribution is the maximal entropy distribution which retains finite mean and variance, as well as being computationally convenient.].
Three-layer models with Gaussian latent fields comprise a wide array of popular models, including generalised linear mixed models (GLMMs), and have been collectively studied under the title latent Gaussian models (LGMs) [@rue2009approximate].

@@

Decomposing the latent field in a LGM gives the sum of a constant mean $\bm{\mu}$, known also as the fixed effects, and (spatial) random effects $\bphi \sim \mathcal{N}(\mathbf{0}, \mathbf{\Sigma})$.
The mean may be modelled using a linear predictor $\mu_i = \beta_0 + \z_i^\top \bm{\beta}$, where $(\beta_0, \bm{\beta})$ are regression parameters corresponding to areal covariates^[To denote length $p$ vectors of areal and point covariates, we use $\z_i = (z_{i,1}, \ldots, z_{i,p})^\top$ and $\z(s) = (z_1(s), \ldots, z_p(s))^\top$ respectively.} $(1, \z_i)^\top$].
Our focus is on the spatial random effects $\bphi$, whose role is to capture the spatial structure between areas not already accounted for in the mean.
Specifying $\bphi$ amounts to specifying the entries of a precision or covariance matrix, which we discuss below.

If no spatial structure remains after inclusion of covariates in the mean, independent and identically distributed (IID) random effects $\phi_i \sim \N (0, \tau_\phi^{-1})$ should be used, where $\tau_\phi$ is a shared precision.
Exploratory data analysis techniques such as Moran's $I$ coefficient [@cliff1981spatial], visual inspection, may be used to determine if there remains any spatial auto-correlation in the data.
<!-- TODO: Add calculation of Moran's $I$ to my HIV survey data-sets. -->

@@

\begin{figure}
\centering
\includegraphics{depends/geometry-graph-zwe.pdf}
\caption{A map of the districts of Zimbabwe together with the corresponding adjacency graph structure $\mathcal{G}$.}
\label{fig:zi}
\end{figure}

@@

One way to encode spatial structure between areas is by using a symmetric relation, where $i \sim j$ if the areas $A_i$ and $A_j$ are adjacent or neighbouring, as often defined by a shared border^[Any choice of adjacency definition is possible. For example, one might chose to allow inclusion of second degree neighbours resulting in so-called higher-order models [@paciorek2013spatial].].
The Besag model [@besag1991bayesian] is a type of improper conditional auto-regressive (ICAR) model where the full conditional distribution of the $i$th spatial random effect is given by
\begin{equation}
    \phi_i \, | \, \bphi_{-i} \sim \N \left(\frac{1}{n_{\delta i}} \sum_{j: j \sim i} \phi_j, \frac{1}{n_{\delta i}\tau_\phi}\right), \label{eq:besag}
\end{equation}
where $\delta i$ is the set of neighbours of $A_i$ with cardinality $n_{\delta i} = |\delta i|$ and $\bphi_{-i} = (\phi_1, \ldots, \phi_{i - 1}, \phi_{i + 1}, \ldots, \phi_n)^\top$ is the vector of spatial random effects with the $i$th entry removed.
The conditional mean of the random effect $\phi_i$ is the average of its neighbours $\{\phi_j\}_{j \sim i}$ and the precision $n_{\delta i}\tau_\phi$ is proportional to the number of neighbours $n_{\delta i}$.
By Brook's lemma \cite[Chapter 2]{rue2005gaussian} the set of full conditionals of the Besag model are equivalent to the Gaussian Markov random field (GMRF) given by
\begin{equation}
    \bphi \sim \mathcal{N}(\mathbf{0}, \tau_\phi^{-1} \mathbf{R}^{-}), \label{eq:gmrf}
\end{equation}
where $\mathbf{R}^{-}$ is the generalised inverse of the rank-deficient structure matrix\footnote{So-called because it defines the \textit{structure} of the precision matrix $\mathbf{Q} = \tau_\phi \mathbf{R}$.} $\mathbf{R}$ with entries
\begin{equation}
    R_{ij} =
    \begin{cases}
        n_{\delta i}, & i = j \\
        -1, & i \sim j \\
        0, & \text{otherwise.}
    \end{cases}
\end{equation}
The Markov property arises due to the conditional independence structure $p(\phi_i \, | \, \bphi_{-i}) = p(\phi_i \, | \, \bphi_{\delta i})$ whereby each area only depends on its neighbours.
This is reflected in the sparsity of $\mathbf{R}$ whereby $\phi_i \perp \phi_j \, | \, \phi_{-ij}$ if and only if $R_{ij} = 0$.
The structure matrix $\mathbf{R}$ may also be expressed as the Laplacian of the adjacency graph $\mathcal{G} = (\mathcal{V}, \mathcal{E})$ with vertices $v \in \mathcal{V}$ corresponding to each area and edges $e \in \mathcal{E}$ between vertices $i$ and $j$ when $i \sim j$.
Figure \ref{fig:zi} shows the adjacency graph for the districts of Zimbabwe alongside the geometry.
Shortly, in Section \ref{sec:concerns} we discuss the appropriateness of using the graph rather than the full geometry.

Rewriting Equation \ref{eq:gmrf}, the probability density function of $\bphi$ is
\begin{equation}
    p(\bphi)
    \propto \exp \left( -\frac{\tau_\phi}{2} \bphi^\top \mathbf{R} \bphi \right)
    \propto \exp \left( -\frac{\tau_\phi}{2} \sum_{i \sim j} (\phi_i - \phi_j)^2 \right). \label{eq:pdfphi}
\end{equation}
This density is a function of the pairwise differences $\phi_i - \phi_j$ and so is invariant to the addition of a constant $c$ to each entry $p(\bphi) = p(\bphi + c\mathbf{1})$, leading to an improper uniform distribution on the average of the $\phi_i$.
If $\mathcal{G}$ is connected\footnote{In that by traversing the edges, any vertex can be reached from any other vertex.} then there is only one impropriety in the model and $\text{rank}(\mathbf{R}) = n - 1$, while if $\mathcal{G}$ is disconnected, and composed of $n_c \geq 2$ connected components with index sets $I_1, \ldots, I_{n_c}$, then the corresponding structure matrix $\mathbf{R}$ has rank $n - n_c$ and the density is invariant to the addition of a constant to each of the connected components $p(\bphi_{I}) = p(\bphi_{I} + c \mathbf{1})$ where $I = I_1, \ldots, I_{n_c}$.

<!-- ## Concerns about the Besag model's representation of space \label{sec:concerns} -->

<!-- \begin{figure} -->
<!-- \centering -->
<!-- \includegraphics{depends/maup1.pdf} -->
<!-- \caption{Each of these four geometries corresponds to the same adjacency graph.} -->
<!-- \label{fig:tikz1} -->
<!-- \end{figure} -->

<!-- \begin{figure} -->
<!-- \centering -->
<!-- \includegraphics{depends/maup2.pdf} -->
<!-- \caption{A sequence of geometries where the number of neighbours of area one grows by one at each iteration.} -->
<!-- \label{fig:tikz2} -->
<!-- \end{figure} -->

<!-- \begin{figure} -->
<!-- \centering -->
<!-- \includegraphics{depends/maup3.pdf} -->
<!-- \caption{Each of the shaded areas are split into two moving from geometry (i) to (ii).} -->
<!-- \label{fig:tikz3} -->
<!-- \end{figure} -->

<!-- The Besag model was originally proposed for use in image analysis, where areas correspond to pixels arranged in a regular lattice structure. -->
<!-- Since then, it has seen wider use, including in situations where the spatial structure is less regular. -->
<!-- This leads to a number of concerns about the model's applicability, which we highlight below. -->
<!-- Closely linked are the modifiable areal unit problem [@openshow1979million] whereby statistical conclusions change as a result of seemingly arbitrary changes in data aggregation, and the challenge of ecological inference and the ecological fallacy [@wakefield2010aggregation], which applies particularly to inference about individuals and groups. -->
<!-- <!-- Missing an item about potential loss of information by using a sparse matrix, though I don't actually think this is very important. --> -->

<!-- ### Adjacency compression -->

<!-- Summarising all spatial information with an adjacency graph represents a loss of information. -->
<!-- Many geometries share the same adjacency graph, and therefore the same probability model, as illustrated by Figure \ref{fig:tikz1}. -->
<!-- This fact is not concerning in itself, though it prompts consideration as to whether the class of geometries with the same adjacency graph is sufficiently similar to merit identical models.\footnote{The regularity of realistic geometries may help to constrain each class to be more similar than it strictly has to be. In other words, although pathological geometries can be constructed, they are implausible in statistical practise and so not of great concern to us here.} -->
<!-- It is intuitive that the more regular the spatial structure the less information is lost in compression to an adjacency graph. -->
<!-- In image analysis, very little spatial information is lost in compression a lattice structure to an adjacency graph. -->
<!-- On the other hand, the regions of a country, determined by political and geographic forces, tend to display greater irregularity. -->
<!-- As such, the appropriateness of adjacency compression varies by the type of geometry common to the application setting. -->

<!-- ### Mean structure -->

>>>>

---
title: Models for spatial structure in small-area estimation
author:
  - name: Adam Howes
    email: ath19@ic.ac.uk
    affiliation: A
    corresponding: ath19@ic.ac.uk
  - name: Harrison Zhu
    affiliation: A
  - name: Yidan Xu
    affiliation: C
  - name: Samir Bhatt
    affiliation: D
  - name: Jeffrey W. Eaton
    affiliation: B
  - name: Seth R. Flaxman
    affiliation: E
address:
  - code: A
    address: Department of Mathematics, Imperial College London
  - code: B
    address: MRC Centre for Global Infectious Disease Analysis, Imperial College London
  - code: C
    address: Department of Statistics, University of Michigan
  - code: D
    address: Department of Public Health, University of Copenhagen
  - code: E
    address: Department of Computer Science, University of Oxford
abstract: |
  **Background:**

@@

```{r echo = FALSE}
options(scipen = 100)
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  dpi = 320,
  cache = TRUE,
  out.width = "95%",
  fig.align = 'center'
)

comma <- function(x) formatC(x, format = "d", big.mark = ",")
```

@@

Spatial random effects are frequently used to model spatial variation in areal data [@haining2003spatial; @cramb2018investigation].
Gaussian Markov random fields (GMRFs) [@rue2005gaussian], which combine a Gaussian distribution with Markov conditional independence assumptions between areas, are the most common class of models used to specify spatial random effects.
Using a GMRF typically reflects the assumption that observations made in areas close together are correlated, and more distant relationships may be ignored.
The simplest GMRF model is arguably that of @besag1991bayesian, where information is borrowed equally from each adjacent area.
This model requires minimal additional modelling choices and is widely implemented, including as a part of the `R-INLA` package [@rue2017bayesian], popular within spatial statistics.
Use of this model is widespread, including in agriculture [@oliver2015soil], ecology [@saracco2010modeling], epidemiology [@lawson2013statistical], image analysis [@schmid2006bayesian], neuroscience [@gossl2001bayesian] and public health [@dwyer2015drinking].
However, for irregular geometries like the administrative divisions of a country, the assumptions made about space are unrealistic.
We hypothesised that incorporating a spatial structure more faithfully to the specific geometry would improve the performance of small-area estimation models.
Here, we set out to test this hypothesis, and in doing so provide practical recommendations for modelling spatial structure.
<!-- @gutreuter2019improving use the SAR model, which is still adjacency based but is not exactly the Besag model--restructure introduction to be more general and include introduction of other adjacency based models (which all have the same criticisms)? -->

<!-- Within the GRMF framework, attempts to more accurately reflect irregular spatial structure have defined weights specifying the extent to which neighbours are related. -->
<!-- @duncan2017spatial compared seventeen methods for specifying these weights, but surprisingly did not find any which outperform the Besag model^[This conclusion was specific to the (often-analysed) Scottish lip cancer example, and like the majority of the literature, based on the deviance information criteria (DIC). The DIC is based on summarising the posterior distribution by a point estimate, and is recommended against by @vehtari2017practical.]. -->
<!-- Another approach is to take into account that areal data is typically produced by aggregating higher resolution data, so as to be a particular instance of the broader challenge of inference about a variable at a different resolution to that it was observed at, known in geostatistics as the change-of-support problem [@gelfand2001change] and dating to foundational work on block kriging [@krige1951statistical; @matheron1976forecasting]. -->
<!-- Ideas from change-of-support modelling are applied by @kelsall2002modeling to consider areal data as coming about by an aggregation of a continuously-indexed Gaussian process, resulting in a covariance structure between two areas given by the average covariance between two points chosen randomly from those areas. -->
<!-- This type of model is particularly natural in the log-Gaussian Cox process modelling framework [@li2012log; @diggle2013spatial; @taylor2015bayesian; @johnson2019spatially], `lgcp` [@taylor2015bayesian] and `SDALGCP` [@johnson2019spatially] packages. -->
<!-- Inference for aggregated data has recently been advanced by @wilson2018pointless who consider the SPDE approach of @lindgren2011explicit, using `R-INLA`, and an empirical Bayes approach, using `TMB` [@kristensen2015tmb], and made available as a part of the `disaggregation` package [@nandi2020disaggregation]. -->

<!-- We also hypothesised that improvements to the Besag model may be transferable to improving convolution models. -->
<!-- Recognising that spatial heterogeneity in outcomes commonly reflects a combination of both spatially correlated processes and specific circumstances at a given location, it is usually recommended to use a random effect specification which includes both spatially structured and unstructured components. -->
<!-- Examples include the BYM2 model [@simpson2017penalising] and earlier convolutions such as the BYM [@besag1991bayesian], Leroux [@leroux2000estimation] or Dean [@dean2001detecting] models. -->
<!-- Evidence for this recommendation includes the comparison studies of @lee2011comparison and @riebler2016intuitive. -->
<!-- These convolutions use a Besag model as the spatially structured component, though any choice of spatial random effect model is possible. -->
<!-- The Naomi HIV small-area estimation model [@eaton2021naomi], which motivates our work, uses both Besag and BYM2 spatial random effects, at different levels of the hierarchical model, and more broadly both models are of substantive practical interest. -->

@@

The remainder of this paper is organised as follows.
Section \ref{sec:background} provides background on areal data and the Bayesian hierarchical modelling approach to small-area estimation.
In Section \ref{sec:ab}, we review developments in specifying spatial random effects based on adjacency, before presenting an alternative approach based on kernels in Section \ref{sec:ks}.
In Section \ref{sec:simulation} we compare models using simulated data, before applying them to mapping HIV prevalence in sub-Saharan Africa in Section \ref{sec:hiv}.
Finally, we discuss our conclusions and directions for future research in Section \ref{sec:conclusion}.

@@

Let $\Sc \subset \mathbb{R}^2$ be the study region, and the disjoint areas $\{A_i\}_{i = 1}^n$ be a spatial partition of $\Sc$.
Areal data are a type of spatial data where observations $y_i$ are associated to areas $A_i$.
Examples include the colour of a pixel, the minimum wage in a state, and the number of disease cases in a region.
Point data are another type of spatial data where instead observations $y(s)$ can be made at any location $s \in \Sc$ of a spatially-continuous stochastic process.

Usually, areal may be conceptualised as arising from aggregation of point data so that $y_i = \int_{A_i} y(s) \text{d}s$.
Exceptions include policies determined at an administrative level, such as the minimum wage or disease control measures, which have a genuinely discrete spatial structure\footnote{For data of this sort, there is less clear justification for using spatial smoothing. Realistically, data relate to many covariates, some number of which are likely have discrete spatial structure and others of which continuous spatial structure, such that the result is often an intermediate. It may be possible to model this explicitly, though the result might look similar to what is achieved using e.g. BYM2.}.
If the areal data we observe are indeed an aggregation of point data, why not use the higher resolution point data in our models rather than areal data?
Explainations include that (a) the underlying point data are unobserved, (b) although point data are collected, they may not be accessible due to privacy constraints, administrative practicality or storage capacity, and (c) the point data are available, but we decide not to use them in the model.
In the final case (c), researchers may prefer to use area-level models because models for point data are typically more complex, require more data, including covariate data which may not exist, and have higher computational burden.
Choosing to model the underlying data generating mechanism as either areal or point is often a matter of pragmatism.

@@

The latter benefit is basis for the statistical task of small-area estimation, which aims to produce reliable local estimates where small sample sizes lead to noisy data [@pfeffermann2013new].
In a spatial setting, this is often in small geographic areas, though the phrase "small-area" is not restricted to geographic areas and may be interpreted more broadly to mean any area where data are insufficient to make accurate local parameter inferences.
In the context of multilevel regression and post-stratification [@gelman1997poststratification], small-areas are generated by the intersection of demographic variables like age, gender and race, alongside geographic variables.
Due to the cost of gathering samples, a survey may be designed to give reliable estimates at at an aggregated level but not at a small-area level.
Although direct estimators of local parameters are unbiased, when data are sparse the total error may be reduced by accepting some bias in exchange for reduced variance using so-called indirect estimators.
Smoothing approaches use information from similar units to "borrow strength" from one parameter to another, with determining precisely what is meant by "similar" a central challenge.
@wakefield2020small review design-based approaches and model-based approaches to spatial analysis of health-indicators, including both area and unit-level analysis.

@@

Bayesian hierarchical models are an attractive framework for small-area estimation, whereby areal data $y$ corresponding to the areas $\{A_i\}_{i = 1}^n$, are modelled using a three-layer structure [@berliner1996hierarchical; @cressie2015statistics; @rao2015small] given by
\begin{alignat}{2}
&\text{(Observations)}  &        y_i &\sim p(y_i \, | \, g^{-1}(x_i), \theta), \quad i = 1, \ldots, n, \label{eq:data} \\
&\text{(Latent field)}  &        x &\sim p(x \, | \, \theta), \label{eq:process} \\
&\text{(Parameters)}    & \qquad \theta &\sim p(\theta), \label{eq:parameters}
\end{alignat}
where $x$ is the $n$-dimensional latent field, $\theta$ are the $m$-dimensional parameters, $m < n$, and $g$ is a link function.
Spatial structure is modelled at the level of the latent field, and the observation model, often either binomial or Poisson, is conditionally independent and identically distributed $p(\y \, | \, x, \theta) = \prod_{i = 1}^n p(y_i \, | \, x_i, \theta)$.
As small-area estimation models are descriptive rather than mechanistic, it is natural [@best2005comparison] to use a flexible and computationally efficient Gaussian distribution $x \sim \mathcal{N}(\mu, \Sigma)$ for the latent field.
Three-layer models with Gaussian latent fields comprise a wide array of popular models, including generalised linear mixed models (GLMMs), and have been collectively studied under the title latent Gaussian models (LGMs) [@rue2009approximate].

@@

The latent field in a LGM is comprised a constant mean $\mu$, known also as the fixed effects, and (spatial) random effects $\phi \sim \mathcal{N}(0, \Sigma)$.
The mean may be modelled using a linear predictor $\mu_i = \beta_0 + z_i^\top \beta$, where $(\beta_0, \beta)$ are regression parameters corresponding to areal covariates^[To denote length $p$ vectors of areal and point covariates, we use $z_i = (z_{i,1}, \ldots, z_{i,p})^\top$ and $z(s) = (z_1(s), \ldots, z_p(s))^\top$ respectively.} $(1, \z_i)^\top$].
Here, we focus on the spatial random effects $\phi$, whose role is to capture the spatial structure between areas not already accounted for in the mean.
If no spatial structure remains after inclusion of covariates in the mean, independent and identically distributed (IID) random effects $\phi_i \sim \N (0, \tau_\phi^{-1})$ should be used, where $\tau_\phi$ is a shared precision.
Exploratory data analysis techniques such as visual inspection or Moran's $I$ coefficient [@cliff1981spatial] may be used to determine if there remains any spatial auto-correlation in the data.
<!-- TODO: Add calculation of Moran's $I$ to my HIV survey data-sets. -->
Specifying $\phi$ amounts to specifying the entries of a precision or covariance matrix, as we discuss below.

@@

```{r geometry-graph-zwe, fig.align="center", fig.cap="A map of the districts of Zimbabwe together with the corresponding adjacency graph structure $\\mathcal{G}$. \\label{fig:geometry-graph-zwe}"}
knitr::include_graphics("depends/geometry-graph-zwe.pdf")
```

@@

One way to encode spatial structure between areas is by using a symmetric relation, where $i \sim j$ if the areas $A_i$ and $A_j$ are adjacent or neighbouring, as often defined by a shared border^[Any choice of adjacency definition is possible. For example, one might chose to allow inclusion of second degree neighbours resulting in so-called higher-order models [@paciorek2013spatial].].
The Besag model [@besag1991bayesian] is a type of improper conditional auto-regressive (ICAR) model where the full conditional distribution of the $i$th spatial random effect is given by
\begin{equation}
    \phi_i \, | \, \phi_{-i} \sim \N \left(\frac{1}{n_{\delta i}} \sum_{j: j \sim i} \phi_j, \frac{1}{n_{\delta i}\tau_\phi}\right), \label{eq:besag}
\end{equation}
where $\delta i$ is the set of neighbours of $A_i$ with cardinality $n_{\delta i} = |\delta i|$ and $\phi_{-i} = (\phi_1, \ldots, \phi_{i - 1}, \phi_{i + 1}, \ldots, \phi_n)^\top$ is the vector of spatial random effects with the $i$th entry removed.
The conditional mean of the random effect $\phi_i$ is the average of its neighbours $\{\phi_j\}_{j \sim i}$ and the precision $n_{\delta i}\tau_\phi$ is proportional to the number of neighbours $n_{\delta i}$.
By Brook's lemma \cite[Chapter 2]{rue2005gaussian} the set of full conditionals of the Besag model are equivalent to the Gaussian Markov random field (GMRF) given by
\begin{equation}
    \phi \sim \mathcal{N}(\mathbf{0}, \tau_\phi^{-1} R^{-}), \label{eq:gmrf}
\end{equation}
where $R^{-}$ is the generalised inverse of the rank-deficient structure matrix\footnote{So-called because it defines the \textit{structure} of the precision matrix $\mathbf{Q} = \tau_\phi R$.} $R$ with entries
\begin{equation}
    R_{ij} =
    \begin{cases}
        n_{\delta i}, & i = j \\
        -1, & i \sim j \\
        0, & \text{otherwise.}
    \end{cases}
\end{equation}
The Markov property arises due to the conditional independence structure $p(\phi_i \, | \, \phi_{-i}) = p(\phi_i \, | \, \phi_{\delta i})$ whereby each area only depends on its neighbours.
This is reflected in the sparsity of $R$ whereby $\phi_i \perp \phi_j \, | \, \phi_{-ij}$ if and only if $R_{ij} = 0$.
The structure matrix $R$ may also be expressed as the Laplacian of the adjacency graph $\mathcal{G} = (\mathcal{V}, \mathcal{E})$ with vertices $v \in \mathcal{V}$ corresponding to each area and edges $e \in \mathcal{E}$ between vertices $i$ and $j$ when $i \sim j$.
Figure \ref{fig:zi} shows the adjacency graph for the districts of Zimbabwe alongside the geometry.
Shortly, in Section \ref{sec:concerns} we discuss the appropriateness of using the graph rather than the full geometry.

Rewriting Equation \ref{eq:gmrf}, the probability density function of $\phi$ is
\begin{equation}
    p(\phi)
    \propto \exp \left( -\frac{\tau_\phi}{2} \phi^\top R \phi \right)
    \propto \exp \left( -\frac{\tau_\phi}{2} \sum_{i \sim j} (\phi_i - \phi_j)^2 \right). \label{eq:pdfphi}
\end{equation}
This density is a function of the pairwise differences $\phi_i - \phi_j$ and so is invariant to the addition of a constant $c$ to each entry $p(\phi) = p(\phi + c\mathbf{1})$, leading to an improper uniform distribution on the average of the $\phi_i$.
If $\mathcal{G}$ is connected\footnote{In that by traversing the edges, any vertex can be reached from any other vertex.} then there is only one impropriety in the model and $\text{rank}(R) = n - 1$, while if $\mathcal{G}$ is disconnected, and composed of $n_c \geq 2$ connected components with index sets $I_1, \ldots, I_{n_c}$, then the corresponding structure matrix $R$ has rank $n - n_c$ and the density is invariant to the addition of a constant to each of the connected components $p(\phi_{I}) = p(\phi_{I} + c \mathbf{1})$ where $I = I_1, \ldots, I_{n_c}$.

## Concerns about the Besag model's representation of space \label{sec:concerns}

\begin{figure}
\centering
\includegraphics{depends/maup1.pdf}
\caption{Each of these four geometries corresponds to the same adjacency graph.}
\label{fig:tikz1}
\end{figure}

\begin{figure}
\centering
\includegraphics{depends/maup2.pdf}
\caption{A sequence of geometries where the number of neighbours of area one grows by one at each iteration.}
\label{fig:tikz2}
\end{figure}

\begin{figure}
\centering
\includegraphics{depends/maup3.pdf}
\caption{Each of the shaded areas are split into two moving from geometry (i) to (ii).}
\label{fig:tikz3}
\end{figure}

The Besag model was originally proposed for use in image analysis, where areas correspond to pixels arranged in a regular lattice structure.
Since then, it has seen wider use, including in situations where the spatial structure is less regular.
This leads to a number of concerns about the model's applicability, which we highlight below.
Closely linked are the modifiable areal unit problem [@openshow1979million] whereby statistical conclusions change as a result of seemingly arbitrary changes in data aggregation, and the challenge of ecological inference and the ecological fallacy [@wakefield2010aggregation], which applies particularly to inference about individuals and groups.
<!-- Missing an item about potential loss of information by using a sparse matrix, though I don't actually think this is very important. -->

### Adjacency compression

Summarising all spatial information with an adjacency graph represents a loss of information.
Many geometries share the same adjacency graph, and therefore the same probability model, as illustrated by Figure \ref{fig:tikz1}.
This fact is not concerning in itself, though it prompts consideration as to whether the class of geometries with the same adjacency graph is sufficiently similar to merit identical models.\footnote{The regularity of realistic geometries may help to constrain each class to be more similar than it strictly has to be. In other words, although pathological geometries can be constructed, they are implausible in statistical practise and so not of great concern to us here.}
It is intuitive that the more regular the spatial structure the less information is lost in compression to an adjacency graph.
In image analysis, very little spatial information is lost in compression a lattice structure to an adjacency graph.
On the other hand, the regions of a country, determined by political and geographic forces, tend to display greater irregularity.
As such, the appropriateness of adjacency compression varies by the type of geometry common to the application setting.

### Mean structure

A feature specific to the Besag model is that all adjacent areas count equally.
This assumption is unsatisfying: for most geometries, we expect some areas to be more highly correlated to some neighbours than to others.
Figure \ref{fig:tikz1} illustrates a number of heuristic features for neighbour importance, including length of shared border in geometry (iii), and the proximity of centers of mass in geometry (iv).
This may be remedied using more general weighted ICAR models, as we outline in Section \ref{sec:icar}.

### Variance structure
